### 2023-03-27 paper-thesis/chapters/24-backend.tex
More changes to the text
More writing work from today

Finally, inductive data types are a bit harder to compile, as Clojure does not have an equivalent to sum types on which we can match. \todo{finish this after implementing}
- atoms voor constructors

>>>>

Finally, inductive data types are a bit harder to compile, as Clojure does not have an equivalent to sum types on which we can match. Instead, we compile inductive datatypes to a pair (constructor, list of constructor arguments).

\todo{finish this after implementing}
- atoms voor constructors

The implementation of the language is now complete, both the type checker and the back-end. We will now discuss editor services (chapter \ref{chap:editor-services}) and compare the implementation with implementations in Haskell (chapter \ref{ch:comp-haskell}) and LambdaPi (chapter \ref{ch:comp-lambdapi}).

### 2023-03-27 paper-thesis/chapters/30-comparing-with-haskell.tex
More writing work from today

In this chapter, we implement the language defined in chapter~\ref{chap:baselang} in Haskell, and then compare the implementation with the implementation in Statix. We want the design of the implementation in Haskell to be similar to conventional implementations of dependently typed languages, so we can compare the Statix implementation with them.

\section{Defining the AST}
To implement the calculus of constructions in Haskell, we first define the \verb|Expr| datatype. We chose to define the language using De Bruijn indices, since this is the convention when implementing dependently typed languages, and the goal of this implementation in Haskell is to compare the Statix implementation with conventional ones. 
\begin{lstlisting}
data Expr =
	Type
	| Let Expr Expr
	| Var Int
	| FnType Expr Expr
	| FnConstruct Expr Expr
	| FnDestruct Expr Expr
\end{lstlisting}

@@

\begin{lstlisting}
NType(Var(0)) :: NType(Type()) :: Nil
\end{lstlisting}
\begin{tikzpicture}[scopegraph, node distance = 2.5em and 2.5em]
	\node[scope] (s1) {$s1$};
	\node[type, above of = s1] (s1l) {$\texttt{T : Type()}$};
	\draw (s1) edge[type] (s1l);

@@

	\draw (s2) edge[] (s1);
\end{tikzpicture}

@@

@@

Now we want to define beta head reduction. Remember that in Statix, beta head reduction is a relation with the signature
\begin{lstlisting}
betaReduceHead : 
	(scope * Expr) * list((scope * Expr)) -> (scope * Expr)
\end{lstlisting}

@@

>>>>

In this chapter, we implement the language defined in chapter~\ref{chap:baselang} in Haskell, and then compare the implementation with the implementation in Statix. We want the design of the implementation in Haskell to be representative of conventional implementations of dependently typed languages, so we can use it to compare the Statix implementation with them.

\section{Defining the AST}
To implement the calculus of constructions in Haskell, we first define the inductive datatype \verb|Expr|. We chose to define the language using De Bruijn indices, since this is the convention when implementing dependently typed languages, and the goal of this implementation in Haskell is to compare the Statix implementation with conventional ones. 
\begin{lstlisting}
data Expr =
	Type
	| Let Expr Expr
	| Var Int
	| FnType Expr Expr
	| FnConstruct Expr Expr
	| FnDestruct Expr Expr
\end{lstlisting}
There is a constructor for each constructor in our language. The difference with the implementation in Spoofax is that the \verb|Let|, \verb|FnType|, and \verb|FnConstruct| no longer store an identifier, instead they introduce a new binding into their body, which are accessible by their De Bruijn indices.

@@

\begin{lstlisting}
NType(Var(0)) :: NType(Type()) :: Nil
\end{lstlisting}
\begin{tikzpicture}[scopegraph, node distance = 2.5em and 2.5em]
	\node[scope] (s0) {$s0$};

	\node[scope, right of = s0, xshift=5em] (s1) {$s1$};
	\node[type, above of = s1] (s1l) {$\texttt{T : Type()}$};
	\draw (s1) edge[type] (s1l);

@@

	\draw (s1) edge[] (s0);
	\draw (s2) edge[] (s1);
\end{tikzpicture}

@@

\section{Defining Type Checking}

Next we need to define type checking. We define a function \verb|tc|:
\begin{lstlisting}
tc :: SExpr -> Either String Expr
\end{lstlisting}

The function returns a \verb|String| in case of an error, or a \verb|Expr| which is the type of the expression. This is different than in Statix, where errors are automatically created during type checking by Statix. We use the power of the \verb|Either String| monad to our advantage, so we can define our function using \emph{do blocks}. For example, compare the implementation of \verb|Let| in figure \ref{fig:type-check-rules} (repeated below for convenience) with the implementation in Haskell:

\begin{lstlisting}
tc (s, Let e b) = do
	_ <- tc (s, e)
	bt <- tc (sPutSubst s (s, e), b)
	pure (shift (-1) 0 bt)
\end{lstlisting}

\begin{mathpar}
		\inferrule{
	\toe{s}{e}{t_e}
	\\ \toe{\co{sPutSubst}(s, x, (s, e))}{b}{t_b}
}{
	\toe{s}{\co{Let}(x, e, b)}{t_b}
}
\end{mathpar}

We are using do blocks to pass errors through the type checking function. The only significant difference with the inference rule is that we need to shift the resulting type by one, since the binding of the let-bound variable does not exist at the type level. This shifting is required for variables too, other than this the inference rules are converted directly to cases of the Haskell function.

@@

During type checking evaluation may be needed, so we need to define beta reduction. Remember that in Statix, beta head reduction is a relation with the signature:
\begin{lstlisting}
betaReduceHead : 
	(scope * Expr) * list((scope * Expr)) -> (scope * Expr)
\end{lstlisting}

@@

Now we can define beta head reduction as cases of \verb|brh_|, returning \verb|Nothing| if the expression cannot be reduced, in which case \verb|brh| will take the original expression. The cases correspond directly to the cases of figure \ref{fig:beta-head-reduce-rules}.

### 2023-03-27 paper-thesis/chapters/31-comparing-with-lambdapi.tex
More writing work from today

There exist several \emph{logical frameworks}, tools designed specifically for implementing and experimenting with dependent type theories, such as ALF~\cite{MagnussonN93}, Twelf~\cite{PfenningS99}, Dedukti~\cite{BoespflugCH12}, Elf~\cite{pfenning_1991} and Andromeda~\cite{BauerHP20}. Since these tools are designed specifically for the task, implementing the type system takes less effort in them compared to Spoofax, but for other tasks such as defining a parser or editor services they are not as well equipped.

@@

Note that these types can only represent type-correct terms, \emph{intrinsically typed} terms. This is useful because it means the meta language does the type checking for us. The disadvantage is that it requires extra information: We need to give \verb|B| for \verb|FnConstruct| and \verb|A| and \verb|B| for \verb|FnDestruct|, which is information we don't have to provide to type-check terms in Statix. These could easily be automatically generated by a type checker but are tedious to specify manually.

It seems like it should be possible to infer these, but lambdapi fails to infer even the simplest ones. \todo{not sure why}.

@@

We define reduction rules to reduce the symbols we defined in the previous section into lambdapi. Lambdapi can then evaluate the rules using its own semantics, not requiring any additional rules similar to the ones in figure \ref{fig:beta-head-reduce-rules}. The reduction rules are given in figure \ref{fig:lp-rules}. Each of the constructs is reduced to the corresponding construct in the meta language.

>>>>

In chapter \ref{ch:inference} we discussed how to implement inference algorithms, and we concluded in section \ref{analysing-inference} that adding equational unification to Statix would make inference algorithms easier to implement. In this chapter we explore this further, by implementing our dependently typed language in a logical framework.

A logical framework is a language that provides a means to define a type system. There exist several \emph{logical frameworks}, designed specifically for implementing and experimenting with dependent type theories, such as ALF~\cite{MagnussonN93}, Twelf~\cite{PfenningS99}, Dedukti~\cite{BoespflugCH12}, Elf~\cite{pfenning_1991} and Andromeda~\cite{BauerHP20}. Since these tools are designed specifically for the task, implementing the type system takes less effort in them compared to Spoofax, but for other tasks such as defining a parser or editor services they are not as well equipped.

@@

Note that these constructors can only represent type-correct terms, \emph{intrinsically typed} terms. This is useful because it means the meta language does the type checking for us. The disadvantage is that it requires extra information: We need to give \verb|B| for \verb|FnConstruct| and \verb|A| and \verb|B| for \verb|FnDestruct|, which is information we don't have to provide to type-check terms in Statix. These could easily be automatically generated by a type checker but are tedious to specify manually. It seems like it should be possible to infer these, but lambdapi fails to do so. Further research is required to find out why lambdapi does not infer these.

@@

We now define the \emph{reduction rules}, which define how to reduce the language. A rule $a \hookrightarrow b$ states that \verb|a| reduces to \verb|b|. We reduce the constructs in our language to constructors in lambdapi, this is called \emph{embedding} the language. Lambdapi can then evaluate the rules using its own semantics, not requiring any additional rules similar to the ones in figure \ref{fig:beta-head-reduce-rules}. The reduction rules are given in figure \ref{fig:lp-rules}. Each of the constructs is reduced to the corresponding construct in the meta language.
